# Type coercion

```{r, include = FALSE}
par(bg = '#fdf6e3')
```

## Seminar

We often need to change the type of a variable. This can be necessary to clean data or because we add a new level to a factor like in the previous section on amending data frames. We begin by loading the england dataset that we created previously.

```{r}
england <- read.csv(file = "england.csv", sep = ",", header = TRUE)
```

Recall that we could not add "Brittany" as a name to our variable. This was because the variable name is stored as a factor and "Brittany" was a new category. The easiest way to add the name "Brittany" is to convert the name variable into a character variable.

### Coerce a factor to character

Let's check our england dataset's variable types.
.
```{r}
str(england)
```

The variable "name" is a factor variable and needs to be converted to a character variable. We coerce the variable into a different type using the `as.character()` function.

```{r}
england$name <- as.character(england$name)
```

Let's inspect the variable types of our dataset again:

```{r}
str(england)
```






